---
title: "Comparative genomic analysis of Flavobacteriaceae: insights into carbohydrate metabolism, gliding motility and secondary metabolite biosynthesis"
author: "Asimenia Gavriilidou"
date: "February 10, 2020"
output:
html_document: default
pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Functional Profiling

This analysis was done using R (version 3.5.0) and RStudio.

##Comparison of Pfam profiles of all strains across taxonomic groups  

Input table: Relative abundance of Pfam annotations per group 

```{r, eval=FALSE}
#Set working directory
setwd("D:/Phd/FLAVOS/Flavos Paper/Input tables")

#Load all required packages
library(vegan)
library(phyloseq)
library(ggplot2)

pfams_all <- read.csv("Rel_Abund_GeneCount_Pfam_ALL.csv", header = T)
row.names(pfams_all) <- pfams_all$Func_id
pfams_all <- pfams_all[,-c(1,2)]

#Transpose table
t.pfams_all <- t(pfams_all)

##Import metadata-groups
metadata_groups <- read.csv("Metadata_Groups.csv", header = T)
row.names(metadata_groups) <- metadata_groups$Genome_name

#Calculate distances (Bray-Curtis dissimilarity)
diss.pfams_all <- vegdist(t.pfams_all, method = 'bray')

#Permanova
permanova_all <- adonis(diss.pfams_all~ metadata_groups$Groups, data = metadata_groups, permutations = 999, method = "bray")
print(as.data.frame(permanova_all$aov.tab)["metadata_groups$Groups", "Pr(>F)"]) ##0.001

#NMDS using Phyloseq
#set "otu_table" -> pfams are rows and genomes (samples) are columns
pfams_all = otu_table(pfams_all, taxa_are_rows = T)

metadata_groups = sample_data(metadata_groups)

#Create phyloseq object
physeq_all = phyloseq(pfams_all, metadata_groups)

#Ordination
ord_groups <- ordinate(physeq_all, "NMDS", "bray")

#Calculate stress of ordination
ord_groups$stress #[1] 0.0941758

#Create NMDS plot
p = plot_ordination(physeq_all, ord_groups, type = "samples", color = "Groups")
p_ord <- p + scale_color_manual(values = c("#33c633", "#253494", "#f98b8b"), name = "Taxa") + geom_point(size = 3) + theme(axis.line = element_line(colour="grey"), panel.background=element_blank(), text = element_text(size = 20, family = "Arial")) 
NMDS_Pfams_all <- print(p_ord)
```

##Comparison of Pfam profiles of all Flavobacteriaceae strains between different clades and host- and non-host associated strains

Input table: Relative abundance of Pfam annotations per Flavobacteriaceae clade

```{r, eval=FALSE}
#Load all required packages
library(vegan)
library(phyloseq)
library(ggplot2)

pfams1 <- read.csv("Rel_Abund_GeneCount_Pfams_FLAVOS.csv", header = T)
pfams_id1 <- pfams1[,c(1)]
row.names(pfams1) <- pfams_id1
pfams1 <- pfams1[,-c(1,2)]

#Transpose table
t.pfams1 <- t(pfams1)

##Import metadata-Clades
metadata1 <- read.csv("Metadata_FLAVOS_comb.csv", header = T)
genomes1 <- metadata1[,c(1)]
row.names(metadata1) <- genomes1
metadata1 <- metadata1[,-c(1)]

#Calculate distances (Bray-Curtis dissimilarity)
diss.pfams1 <- vegdist(t.pfams1, method = 'bray')

#Permanova
permanova <- adonis(diss.pfams1 ~ metadata1$Clades, data = metadata1, permutations = 999, method = "bray")
print(as.data.frame(permanova$aov.tab)["metadata1$Clades", "Pr(>F)"]) ##0.001

permanova <- adonis(diss.pfams1 ~ metadata1$Association, data = metadata1, permutations = 999, method = "bray")
print(as.data.frame(permanova$aov.tab)["metadata1$Association", "Pr(>F)"]) ##0.593

#NMDS using Phyloseq 
#set "otu_table" -> pfams are rows and genomes (samples) are columns
pfams1 = otu_table(pfams1, taxa_are_rows = T)

#import metadata 
metadata1 <- read.csv("Metadata_FLAVOS_comb.csv", header = T)
genomes1 <- metadata1[,c(1)]
row.names(metadata1) <- genomes1
metadata1 <- metadata1[,-c(1)]
metadata1 = sample_data(metadata1)

#Create phyloseq object
physeq1 = phyloseq(pfams1, metadata1)

#Ordination
ord_clades <- ordinate(physeq1, "NMDS", "bray")

#Calculate stress 
ord_clades$stress ## 0.1646015

#NMDS plot
p1 = plot_ordination(physeq1, ord_clades, type = "samples", color = "Clades")
p_ord1 <- p1 + scale_color_manual(values = c("#8c8c8c", "#756bb1", "#39ae9a", "#f9bbbb")) + geom_point(size = 3) + theme(axis.line = element_line(colour="grey"), panel.background=element_blank(), text = element_text(size = 20, family = "Arial"))
print(p_ord1)
```

##Unique and shared protein families between groups (Flavobacteriaceae, Cyanobacteria and Proteobacteria) and clades (within Flavobacteriaceae) 

The unique and shared Pfams were calculated manually from the **Rel_Abund_GeneCount_Pfam_ALL.csv** table.

```{r, eval=FALSE}
#Venn diagrams
##Load required package
library(VennDiagram)

#Groups
Venn_groups <- VennDiagram::draw.triple.venn(area1 = 3843, area2 = 2645, area3 = 3234, n12 = 2003, n23 = 1838, n13 = 2356, n123 = 1648, category = c("Flavobacteriaceae", "Cyanobacteria", "Proteobacteria"), fill = c("darkblue", "#33c633", "#f94242"), lty = "blank", rotation.degree = 30, cex = 1.2, cat.cex = 1, cat.fontfamily = "Arial", fontfamily = "Arial", cat.dist = c(0.08, 0.05, 0.05), cat.pos = c(-20, 20, 150))

#Clades
Venn_clades <- VennDiagram::draw.quad.venn(area1 = 3551, area2 = 2499, area3 = 2554, area4 = 2415, n12 = 2406, n13 = 2384, n14 = 2338, n23 = 2081, n24 = 2031, n34 = 2018, n123 = 2058, n124 = 2021, n134 = 1999, n234 = 1856, n1234 = 1851, category = c("Marine", "Capnocytophaga", "Flavobacterium", "Tenacibaculum-Polaribacter"), fill = c("#39ae9a", "#8c8c8c", "#756bb1", "#f9bbbb" ), lty = "blank", cex = 1.0, cat.cex = 1.0, cat.fontfamily = "Arial", fontfamily = "Arial", cat.pos = c(-5, 5,-10, -5))
```

To check which Pfams contribute most to the differences between taxonomic groups and within Flavobacteriaceae, we did SIMilarity PERcentage (SIMPER) analysis.

##Most contributing pfams between major taxonomic groups

```{r, eval=FALSE}
#Load required package
library(vegan)

setwd("D:/Phd/FLAVOS/Flavos Paper/Input tables")
pfams_all <- read.csv("Rel_Abund_GeneCount_Pfam_ALL.csv", header = TRUE)
pfams_id <- pfams_all[,c(1)]
row.names(pfams_all) <- pfams_id
pfams_all <- pfams_all[,-c(1,2)]

#Transpose table
pfams.t_all <- t(pfams_all)

#Hellinger transformation
t.pfams_all <- decostand(pfams.t_all, method = "total")

#Remove columns (Pfams) that sum to zero
t.pfams_w_not_col_zeros_all <- t.pfams_all[,colSums(pfams.t_all)!=0]

#Import metadata
metadata_pfams_all <- read.csv("Metadata_Groups.csv", header = T)
groups <- as.character(metadata_pfams_all$Groups)

#SIMPER
data_pfams_all <- cbind(t.pfams_w_not_col_zeros, groups)
simper <- simper(data_pfams_all[,1:5173], group = data_pfams_all$groups, permutations = 999, trace = T)
summary <- summary(simper, ordered = T, digits =3)

##SIMPER does pair-wise comparisons

Flavos_Proteos <- as.data.frame(summary["Flavobacteria_Proteobacteria"])
Flavos_Cyanos <- as.data.frame(summary["Flavobacteria_Cyanobacteria"])
Proteos_Cyanos <- as.data.frame(summary["Proteobacteria_Cyanobacteria"])

##Write output in .csv for further analysis
write.csv(Flavos_Proteos, file="Flavos_Proteos_SIMPER.csv")
write.csv(Flavos_Cyanos, file="Flavos_Cyanos_SIMPER.csv")
write.csv(Proteos_Cyanos, file="Proteos_Cyanos_SIMPER.csv")
```

##Most contributing pfams within Flavobacteriaceae clades

```{r, eval=FALSE}
#Load required package
library(vegan)

setwd("D:/Phd/FLAVOS/Flavos Paper/Input tables")
pfams_flavos <- read.csv("Rel_Abund_GeneCount_Pfams_FLAVOS.csv", header = TRUE)
pfams_id_flavos <- pfams_flavos[,c(1)]
row.names(pfams_flavos) <- pfams_id_flavos
pfams_flavos <- pfams_flavos[,-c(1,2)]

#Transpose table
pfams.t_flavos <- t(pfams_flavos)

#Hellinger transformation
t.pfams_flavos <- decostand(pfams.t_flavos, method = "total")

#Remove columns (Pfams) that sum to zero
t.pfams_w_not_col_zeros_flavos <- t.pfams_flavos[,colSums(pfams.t_flavos)!=0]

#Import metadata
metadata_pfams_flavos <- read.csv("Metadata_Clades.csv", header = T)
clades <- as.character(metadata_pfams_flavos$Clade)


#SIMPER
data_pfams_flavos <- cbind(t.pfams_w_not_col_zeros_flavos, clades)
simper <- simper(data_pfams_flavos[,1:3843], group = data_pfams_flavos$clades, permutations = 999, trace = T)
summary_flavos <- summary(simper, ordered = T, digits = 3)

##SIMPER does pair-wise comparisons
Marine_Flavobacterium <- as.data.frame(summary_flavos["Marine_Flavobacterium"])
Marine_Capnocytophaga <- as.data.frame(summary_flavos["Marine_Capnocytophaga"])
Marine_Tenacibaculum <- as.data.frame(summary_flavos["Marine_Tenacibaculum-Polaribacter"])
Flavobacterium_Capnocytophaga <- as.data.frame(summary_flavos["Flavobacterium_Capnocytophaga"])
Flavobacterium_Tenacibaculum <- as.data.frame(summary_flavos["Flavobacterium_Tenacibaculum-Polaribacter"])
Capnocytophaga_Tenacibaculum <- as.data.frame(summary_flavos["Capnocytophaga_Tenacibaculum-Polaribacter"])

##Write output in .csv for further analysis
write.csv(Marine_Flavobacterium, file = "Marine_Flavobacterium_SIMPER_HT.csv")
write.csv(Marine_Capnocytophaga, file = "Marine_Capnocytophaga_SIMPER_HT.csv")
write.csv(Marine_Tenacibaculum, file = "Marine_Tenacibaculum_SIMPER_HT.csv")
write.csv(Flavobacterium_Capnocytophaga, file = "Flavobacterium_Capnocytophaga_SIMPER_HT.csv")
write.csv(Flavobacterium_Tenacibaculum, file = "Flavobacterium_Tenacibaculum_SIMPER_HT.csv")
write.csv(Capnocytophaga_Tenacibaculum, file = "Capnocytophaga_Tenacibaculum_SIMPER_HT.csv")
```

The results from the pair-wise comparisons were further analysed in Excel. 
Only Pfams with the highest significant contribution (> 0.2%, p < 0.05) to the dissimilarity are shown.

#Carbohydrate metabolism and transport

#Compare frequency of CAZymes between major taxonomic groups and within Flavobacteriaceae
```{r, eval=FALSE}
#Load required package
library(dplyr)

#Between taxa
##Import table
data <- read.csv("Cazymes_Frequency_Taxa.csv", header = TRUE)
colnames(data) <- c("GHs/Mbp", "PLs/Mbp", "CEs/Mbp", "GTs/Mbp", "AAs/Mbp", "CBMs/Mbp", "Taxa")

##Kruskal-Wallis test (a=0.05)
kruskal.test(data$`GHs/Mbp`, data = data, g = data$Taxa) ## p-value = 8.425e-05
kruskal.test(data$`PLs/Mbp`, data = data, g = data$Taxa) ##p-value = 0.001825
kruskal.test(data$`CEs/Mbp`, data = data, g = data$Taxa) ##p-value = 0.0004612
kruskal.test(data$`GTs/Mbp`, data = data, g = data$Taxa) ##p-value = 0.01724
kruskal.test(data$`AAs/Mbp`, data = data, g = data$Taxa) ##p-value = 0.05073
kruskal.test(data$`CBMs/Mbp`, data = data, g = data$Taxa) ##p-value = 4.263e-06

#Between Clades
##Import table
data <- read.csv("Cazymes_Frequency_Clades.csv", header = TRUE)
colnames(data) <- c("GHs/Mbp", "PLs/Mbp", "CEs/Mbp", "GTs/Mbp", "AAs/Mbp", "CBMs/Mbp", "Clades")

##Kruskal-Wallis test (a=0.05)
kruskal.test(data$`GHs/Mbp`, data = data, g = data$Clades) ##  p-value = 0.2372
kruskal.test(data$`PLs/Mbp`, data = data, g = data$Clades) ##p-value = 0.0402
kruskal.test(data$`CEs/Mbp`, data = data, g = data$Clades) ##p-value = 0.436
kruskal.test(data$`GTs/Mbp`, data = data, g = data$Clades) ##p-value = 0.09747
kruskal.test(data$`AAs/Mbp`, data = data, g = data$Clades) ##p-value = 0.5607
kruskal.test(data$`CBMs/Mbp`, data = data, g = data$Clades) ##p-value = 0.05646
```

## CAZymes per Mb in the genomes between different taxonomic groups and within Flavobacteriaceae clades

We used the Output file: **Mean_Rel_Abund_Cazymes.xlsx** from the Genome Annotation section.
Then we created two separate files with the Mean Relative Abundance of the different CAZymes (GHs, PLs, CEs, GTs and CBMs) per taxon and per clade.

```{r, eval=FALSE}
#Barplots
##Load required packages
library(ggplot2)
library(dplyr)
library(ggpubr)

##Per taxon
df_taxa <- read.csv("Barplots_Mean_Taxa.csv", header=TRUE)
p <- ggplot(df_taxa, aes(x=Cazymes, y=Mean, fill=Taxa))+geom_bar(position = position_dodge(), stat = "identity", colour='black')+geom_errorbar(aes(ymin=Mean+SE, ymax=Mean-SE), position = position_dodge(0.9), width=0.25)
p1 <- p+labs(x="CAZymes", y="Mean Relative Abundance")+theme(axis.line = element_line(colour="grey"), panel.background=element_blank(), text=element_text(size = 20, family = "Arial"))+scale_fill_manual(values=c("#33c633", "#253494", "#f94242"))

#Per Clade
df_clades <- read.csv("Barplots_Mean_Clades.csv", header = TRUE)
p2 <- ggplot(df_clades, aes(x=Cazymes, y=Mean, fill=Clades))+geom_bar(position = position_dodge(), stat = "identity", colour='black')+geom_errorbar(aes(ymin=Mean+SE, ymax=Mean-SE), position = position_dodge(0.9), width=0.25)
p3 <- p2+labs(x="CAZymes", y="Mean Relative Abundance")+theme(axis.line = element_line(colour="grey"), panel.background=element_blank(), text=element_text(size = 20, family = "Arial"))+scale_fill_manual(values=c("#8c8c8c", "#756bb1", "#39ae9a", "#f9bbbb"))

final_plot <- ggarrange(p1, p3, nrow=1)
final_plot

###Then you use 'Export' and 'Save as Image' options to get the final plot in the desired dimensions
```

#Gliding Motility and Type 9 Secretion 

We created a presence-absence table with the Gliding motility and T9SS proteins found in the analysed genomes after BLASTp searches.

Input table: **Gld_T9SS_All.csv**

```{r, eval=FALSE}
#Heatmap
##Load required packages
library(devtools)
library(ComplexHeatmap) 
library(circlize) 

#Import table
data <- read.csv("Gld_T9SS_All.csv", header = TRUE)
genomes <- data[,c(1)]
row.names(data) <- genomes
data <- data[,-c(1)]
data <- as.matrix (data)

#Import metadata
metadata <- read.csv("Metadata_Gld_T9SS.csv", header = TRUE)
gliding <- as.character(metadata$Motility)
groups <- as.character(metadata$Groups) 

#Heatmap input table
data2 <- cbind(data, gliding, groups)
data2 <- as.data.frame(data2)
colnames(data2) <-  c("GldA", "GldF", "GldG", "GldB",	"GldD", "GldH", "GldI",	"GldJ",	"GldK", "GldL",	"GldM",	"GldN",	"SprA",	"SprE", 	"SprF",	"SprT", "PorQ",	"PorU", "PorV",	"SprB",	"RemA",	"SigP",	"PorX",	"PorY",	"PorZ",	"SprC",	"SprD", "Motility", "Groups")

#Set colors for the main heatmap
col_fun = colorRamp2(c(0,1), c("white", "#c6dbef"))

#Set colors for annotations
col_motility <- c("blue", "black", "#d93476", "white")
names(col_motility) <- c("gliding", "non-gliding", "other", "unknown")
col_groups <- c("#39ae9a", "#8c8c8c", "#756bb1", "#f9bbbb", "#33c633", "#f94242", "#faf59d")
names(col_groups) <- c("Marine", "Capnocytophaga", "Flavobacterium", "Tenacibaculum-Polaribacter", "Cyanobacteria", "Proteobacteria", "Bacteroidetes")

#Create annotations
ha = rowAnnotation(Motility = anno_simple(data2$Motility, which = "row", gp = gpar(fontsize=10), col = col_motility), annotation_name_side = "bottom", show_annotation_name = TRUE, show_legend = TRUE)
ha1 = rowAnnotation(Groups = anno_simple(data2$Groups, which = "row", gp = gpar(fontsize=10), col = col_groups), annotation_name_side = "bottom", show_annotation_name = TRUE, show_legend = TRUE)

#Create legends for annotations
lgd_ha = Legend(labels = c("T9SS-based gliding", "Non-gliding", "Other", "NA"), title = "Motility", legend_gp =gpar(fill=c("blue", "black", "#d93476", "white")))
lgd_ha1 = Legend(labels = c("Marine", "Capnocytophaga", "Flavobacterium", "Tenacibaculum-Polaribacter", "Cyanobacteria", "Proteobacteria", "Bacteroidetes"), title = "Groups", legend_gp = gpar(fill=c("#39ae9a", "#8c8c8c", "#756bb1", "#f9bbbb", "#33c633", "#f94242", "#faf59d")))

#Create heatmap
ht = Heatmap(data, width = unit(10, "cm"), height = unit(20, "cm"), row_names_gp = gpar (fontsize=10, fontface="italic"), column_names_gp = gpar (fontsize=10), cluster_rows = FALSE, show_column_dend = FALSE, col = col_fun, show_heatmap_legend = FALSE, rect_gp = gpar(col = "white", lwd = 2), column_title_side = "top", cluster_columns = FALSE,left_annotation = ha, right_annotation = ha1)

#Draw heatmap and legends
draw(ht, heatmap_legend_list = list(lgd_ha, lgd_ha1))
```



